\mbox{Specifying Imperative Higher Order Programs}

### 2009-10-01 papers/esop2010/main.tex
go

\begin{abstract}
  %TODO
\end{abstract}

@@

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |- t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

\begin{figure}[htpb]
  \begin{eqnarray*}
    \ceil{α} &=& α\\
    \ceil{τ~\refml_ρ} &=& \ceil{τ}\refml_ρ\\
    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
    \ceil{τ \times τ'} &=& \ceil{τ} \times-> \ceil{τ'}\\
    \ceil{\efft{ε}} &=& \efft{ε}\\
    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)()
  \end{eqnarray*}
  \caption{Lifting types}
  \label{fig:typelift}
\end{figure}

\begin{figure}[htbp]
  \begin{eqnarray*}
    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
    \ceil{\recml~f~(x:τ).\{p\} (t : ε ) \{q\}} &=& 
    (λx:\ceil{τ}.λcur:\efft{ε}.p,λx:\ceil{τ}.λold,cur:\efft{ε}.q)\\
    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
  \end{eqnarray*}
  \caption{Lifting values}
  \label{fig:valuelift}
\end{figure}

We have already seen that programs and specification share in fact the same
syntax. Though they are differently typed, their typing relations still share
a lot. For this reason, and to underline the fact that in Hoare logic, values
are shared between programs and the logic, we present three different typing
relations. One for programs, one for specifications, and the common part is
factored out in a third typing relation for values.

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.~\ref{fig:progtyping}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $τ->τ$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
that the set of terms described by $|-_v$ is precisely the one obtained by
combining syntactic values and pure application in a well-typed way.

In the other case, we have $Γ|- t : τ, ε$, where $t$ is an expression of
type $τ$ and effect $ε$. Impure applications, \letml-expressions and
\ifml-expressions can only be typed in this way. Effects can be produced only
by function application, where the impure arrow $τ'->^ετ$ is involved. In this
case, the effect of the application is precisely the effect $ε$ of the arrow,
because the two subterms are semantic values and don't have any effect.

The typing relies on two predefined types, \propml\ and \boolml. \propml\ is
the logical truth and contains the values \Trueml\ and \Falseml, while
\boolml\ describes boolean (decidable) properties and contains \trueml\ and
\falseml. The distinction between \propml\ and \boolml\ is the same one as in
\coq, for example. Properties expressed in \propml\ are potentially
undecidable and one cannot take decisions by testing their truth value. On the
other hand, \boolml\ talks about decidable properties and so it makes sense to
expect the predicate of the \ifml-statement to be of type \boolml.

missing : 
* subeffecting

%TODO is the LOGIC rule correct ??

\subsection{predefined constants}

arithmetic operations, operations on pairs, logic operations on prop, pre/post

reference operations
!!

operations on maps : restrict, combine
\subsection{Sharing between logic and programs}

use logic application in programs : no problem;
use lambdas with effects in logic : specification

@@

@@

@@

$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$

@@

In the following, we use the metasymbol $v$ to denote well-typed semantic
values, i.e. terms with a typing derivation $Γ|-_v v : τ$.

\begin{figure}[htbp]
  \begin{eqnarray*}
    \wpre_s(v, q) &=& q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}, x:τ.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.\wpre_{s'}(t2, q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
    (\ceil{ t1 } = \trueml => \wpre_s( t2 , q)) /\ \\
    & &(\ceil{ t1 } = \falseml => \wpre_s( t3 , q)) 
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

@@

A prototype implementation by the name ``\who'' exists and is freely available. It implements the
described language with type, region and effect polymorphism, higher-order
logics and store types in the logic. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.

>>>>

% ICFP
%  We present a weakest precondition calculus for an imperative,
%  higher-order programming language, restricted to non-aliasing global
%  references. The main feature of the calculus is {\em effect
%    polymorphism}, {\em i.e.}, higher-order functions can be specified by
%  generalizing over the effects and specifications of their functional
%  arguments. This allows truly modular reasoning about effectful
%  higher-order functions. We achieve this by defining an extension of
%  a standard higher-order logic, more suitable to reason about
%  modifications of the store. The obtained proof obligations, however,
%  are expressible in the standard logic and can be proved using
%  standard theorem provers. We show that the necessary annotations
%  and the obtained proof obligations are quite natural.
%

% Workshop ML
%  We present \who, a tool for verifying effectful higher-order
%  functions. It features {\em Effect polymorphism}, higher-order logic
%  and the possibility to reason about state in the logic, which enable
%  highly modular specifications of generic code. Several small
%  examples and a larger case study demonstrate its usefulness. The
%  \who~tool is intended to be used as an intermediate language for
%  verification tools targeting ML-like programming languages.
%

\begin{abstract} We present a specification language suited for reasoning
  about effectful higher-order programs. The language's main features are
  higher-order logic and effect polymorphism, both of which enable a truly
  modular treatment of higher-order functions. To our knowledge, this is the
  first time that effect polymorphism is put to use in program verification.
  We also define a weakest precondition calculus for this language which
  generates proof obligations in higher-order logic. A simple mechanism based
  on regions permits to restrict the analysis to alias-free programs, thus
  obtaining much simpler proof obligations. Several examples demonstrate the
  practicality of the system. A prototype implementation exists and has been
  used to prove the correctness of these and more examples.  
\end{abstract}

@@

We have already seen that programs and specification share in fact the same
syntax. Though they are differently typed, their typing relations still share
a lot. For this reason, and to underline the fact that in Hoare logic, values
are shared between programs and the logic, we present three different typing
relations. One for programs, one for specifications, and the common part is
factored out in a third typing relation for values. All typing relations
contain an environment $Γ$, which is simply a map from variable names to {\em
type schemes} of the form $∀\alist{χ}.τ$ .

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
  \efft{ε} -> τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

\paragraph{Value typing.}
Let us first look at the common part concerning values. We define a relation
$Γ|-_v t : τ$, which means ``$t$ is well-typed wrt. the environment $Γ$, and
is of type $τ$''. Values cannot have any effect.  Variables and pure
abstractions are typed in the usual way; type instantiations at variable
occurences are substituted in the type. The rule {\sc REC} simply states that
the body of the function can use the argument $x$ as well as the function $f$.
The specifications $p$ and $q$, however, cannot use $f$ and must be well-typed
wrt. the typing relation $|-_l$ described later on. Finally, pure applications
between values are considered as values as well, as they don't have any
effect.

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
  } \and
  { \inferrule*[Left=Sub]
    { Γ |- t : τ', ε \\ ε \subseteq ε' }
    { Γ |- t : τ, ε' }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

\paragraph{Program typing.}
The typing of programs, written $Γ|- t : τ,ε$, assigns a type $τ$ and an
effect $ε$ to an expression $t$, given an environment. This typing is
presented in {\em A-normal form}, a normal form where all intermediate values
are \letml-introduced. All values are also programs with no effect (rule {\sc
Value}). An application must consist of two values and the effect of the
overall expression is precisely the effect of the called function (rule {\sc
App}). There are two rules for \letml, due to the value restriction. One only
can generalize type, effect and region variables when the bound expression is
a value. In this case (rule {\sc LetV}), only the second term may produe
effects. In the other case (rule {\sc Let}), no generalization takes place,
but both terms can have effects. The {\sc Let} rule actually obliges both
terms to have the same effect, but together with the {\sc Sub} rule, which can
increase the effect of an expression, this can always be achieved.
Finally, \ifml-expressions are typed in the usual way, both branches have to
have the same effect.

\paragraph{Logic typing.}

The typing of annotations is basically an extension of the typing for values;
logic terms never have any effect. To ensure this, we need to forbid calling
effectful functions in the logic. One way is to simply omit the typing rule
for effectful function calls. There is, however, another way which also
enables us to inspect the pre/ and postcondition of a function. To this end,
let us introduce the operation of {\em lifting} a type to the logic level (see
Fig.~\ref{fig:typelift}).
\begin{figure}[htpb]
  \begin{eqnarray*}
    \ceil{α} &=& α\\
    \ceil{τ~\refml_ρ} &=& \ceil{τ}~\refml_ρ\\
    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
    \ceil{τ \times τ'} &=& \ceil{τ} \times \ceil{τ'}\\
    \ceil{\efft{ε}} &=& \efft{ε}\\
    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)
  \end{eqnarray*}
  \caption{Lifting types}
  \label{fig:typelift}
\end{figure}
This translation traverses the type and replaces all effectful types by a
tuple type which describes a tuple of a pre- and a postcondition. The
precondition takes as an argument the argument of the function as well as the
part of the store, at the moment of the function call, which corresponds to
the effect of the function. The postcondition's two first arguments are the
same, but in addition it expects the store {\em at the exit of the function},
as well as the function's return value.

\begin{figure}[htbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

Now we can come back to logic typing. Every value is also a logic term,
but now it is assigned its {\em lifted} type. The two other rules are just 
pure variants of their impure counterparts in the program typing.

\paragraph{Predefined constants.}

To actually be able to write programs in this language, we need many
predefined constants and types. The typing relations already mention two,
\propml\ and \boolml. \propml\ is the logical truth and contains the values
\Trueml\ and \Falseml, while \boolml\ describes boolean (decidable) properties
and contains \trueml\ and \falseml. The distinction between \propml\ and
\boolml\ is the same one as in \coq, for example. Properties expressed in
\propml\ are potentially undecidable and one cannot take decisions by testing
their truth value. On the other hand, \boolml\ talks about decidable
properties and so it makes sense to expect the predicate of the
\ifml-statement to be of type \boolml.

The reader may have wondered why the specification language does not contain
any logical connectives. These can of course be declared independently, for
example the conjuction is just a function: \begin{equation*} /\ :  \propml ->
  \propml -> \propml \end{equation*} Actually, when equality is given, all
  logical operations (even quantification) can be defined in the logic.
%TODO citation needed
In the remainder of the paper, we suppose predefined the usual logical
connectives conjunction, disjunction, implication and equivalence, as well as
universal and existential quantification. We also assume all the arithmetic
operations as well as the constructor $(t,t)$  and the accessors $fst$ and
$snd$ for pairs.

% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

The presented programming language, along with the specifications, can, in
fact be used to prove programs. The weakest precondition calculus presented in
the next section can be used to obtain proof obligations which, when proved,
guarantee the correctness of the program with respect to its specification. On
the other hand, in its presented form, and because of the presence of
aliasing, one would constantly need equalities and inequalities between
reference variables to be able to reason about a program.

@@

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the presented typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
to identify memory locations entirely; each region shall describe precisely
one memory location. Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
%TODO

which is only correct if $ ρ1 $ and $ ρ2 $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$. To understand what compatibility means,
consider an {\em atomic} substitution $[ρ|->ρ']$. It is compatible with any
type which does not contain $ρ'$. Similarly, the substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a composite substitution is compatible
%TODO atomic and composite
with a type if each atomic substitution is compatible with the result of the
application of the previous substitution. In this way, it becomes impossible
to have different names for a single reference cell.

The second form of aliasing is when we use the same name for {\em different}
memory locations. This can happen when we allow the user to call $ref$ several
times with the same region instantiation. A simple possibility is to attach
the suppress the $ref$ function entirely and to replace it by a \letml\refml\ 
construct which combines name creation and memory allocation. In this way,
there can of course be only one reference per name. On the other hand, because
of the side condition $ρ\notin τ$, the reference cannot escape the scope of
its \letml\refml\ construct, and so, a function cannot return a reference
which has been created in its body. This restriction could be leveraged a bit if
one seperates name creation and allocation, while still maintaining that one
can only allocate a single reference in each region. This is how it is done in
the correctness proof of this restriction. The presented \letml\refml\
construct is of course a special case.
%TODO now, all operations on maps can be statically resolved.

@@

%TODO
$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$

Subject reduction of our type system is a consequence of subject reduction of
the region calculus of Tofte/Talpin~\cite{tofte97ic}.

@@

We now proceed to explain how proof obligations are obtained from the
annotated program.

In the following, we use the metasymbol $v$ to denote well-typed values, i.e.
terms with a typing derivation $Γ|-_v v : τ$.

\begin{figure}[htbp]
  \begin{eqnarray*}
    \ceil{x~[\alist{κ}]} &=& x~[\alist{κ}]\\
    \ceil{λx:τ.t} &=& λx:τ.\ceil{t}\\
    \ceil{\recml~f~(x:τ).\{p\} t \{q\}} &=& 
    (λx:\ceil{τ}.p,λx:\ceil{τ}.q)\\
    \ceil{t_{τ' -> τ}~t'} &=& \ceil{t}~\ceil{t'}
  \end{eqnarray*}
  \caption{Lifting values}
  \label{fig:valuelift}
\end{figure}

\begin{figure}[htbp]
  \begin{eqnarray*}
    c(x~[\alist{κ}]) &=& \Trueml \\
    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
    c(t~t') &=& c(t) /\ c(t') \\
    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
    ∀s:\efft{ε}.p~s => \wpre_s(t,q \\
    \wpre_s(v, q) &=& q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct(t') /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : τ, ε1 ) ~\inml~ ( t2 : ε2 ) , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε1  ε2 }.λx:\ceil{τ}.\wpre_{s'}(t2, q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

@@

A prototype implementation by the name ``\who''
exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
described language with type, region and effect polymorphism, higher-order
logics and store types in the logic. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.
